---
title: "TidyTuesday Week 24, 2022: Drought Conditions in the US"
author: "Tim Kelso"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE)
```

```{r libraries}
library(tidytuesdayR)
library(extrafont)
library(DataExplorer)
library(tidycensus) #to retrieve FIPS codes
library(tigris) #to download geographic boundaries from US Census Bureau
library(sf)
library(gganimate)
library(gifski)
library(ggforce)
library(lubridate)
library(tidyverse)
```
# Background 
## Data source  
Data from National Integrated Drought Information System (https://www.drought.gov/).  

## Data information  
*drought.csv*
The Standardized Precipitation Index (SPI) is an index to characterize meteorological drought on a range of timescales, ranging from 1 to 72 months, for the lower 48 U.S. states. The SPI is the number of standard deviations that observed cumulative precipitation deviates from the climatological average. NOAA's National Centers for Environmental Information produce the 9-month SPI values below on a monthly basis, going back to 1895.
**(From https://www.drought.gov/historical-information?dataset=1&selectedDateUSDM=20110301&selectedDateSpi=19580901)**

*drought_fips.csv*
The Drought Severity and Coverage Index is an experimental method for converting drought levels from the U.S. Drought Monitor map to a single value for an area. DSCI values are part of the U.S. Drought Monitor data tables. Possible values of the DSCI are from 0 to 500. Zero means that none of the area is abnormally dry or in drought, and 500 means that all of the area is in D4, exceptional drought.

FIPS id: first two digits = state, last 3 digits = county
https://walker-data.com/tidycensus/reference/fips_codes.html

```{r get_data}
# Get the Data
drought_orig <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-06-14/drought.csv')
drought_fips_orig <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-06-14/drought-fips.csv')
```

# Analysis  
## Initial thoughts  
- What state has endured the most/least drought since 2000?  
- Map of average DSCI for each state (heatmap) from 2000 - 2022
- Map of average DSCI for each county (heatmap) from 2000 - 2022
- Scatterplot/violin/density/raincloud plot of average DSCI by month (all counties/states)
- Using additional census data, in which state/county have the most people been affected by drought?  


## Tables/Columns of potential interest:  
- FIPS: separate into state (first two digits) and county (last three digits)    
- DSCI: measure of drought severity in an area  
  


```{r data_transformation}
drought_fips_yearlyAverage_state <- drought_fips_orig %>% 
  mutate(Year = year(date)) %>% 
  group_by(State, Year) %>% 
  summarise(Avg_DSCI = mean(DSCI, na.rm = TRUE)) %>% 
  left_join(fips_codes %>% distinct(state, state_name), by = c("State" = "state"))  #add state full name

drought_fips_monthlyAverage_state <- drought_fips_orig %>% 
  mutate(Month = month(date, label = TRUE)) %>% 
  group_by(State, Month) %>% 
  summarise(Avg_DSCI = mean(DSCI, na.rm = TRUE)) %>% 
  left_join(fips_codes %>% distinct(state, state_name), by = c("State" = "state"))  #add state full name


drought_fips_yearlyAverage_county <- drought_fips_orig %>% 
  mutate(Year = year(date)) %>% 
  group_by(FIPS, Year) %>% 
  summarise(Avg_DSCI = mean(DSCI, na.rm = TRUE)) %>% 
  mutate(FIPS_state = str_extract(FIPS, "^[:digit:]{2}"),
         FIPS_county = str_extract(FIPS, "[:digit:]{3}$")) %>% 
  left_join(fips_codes, by = c("FIPS_state" = "state_code", "FIPS_county" = "county_code"))  #add county full name

#get shapefile for all US states
states <- states(cb = TRUE, class = "sf") %>% 
  filter(!as.numeric(STATEFP) %in% c(2, 15, 60, 66, 69, 72, 78)) # lower 48 only
  
#get shapefile for all US counties
counties <- counties(cb = TRUE, class = "sf") %>% 
  filter(!as.numeric(STATEFP) %in% c(2, 15, 60, 66, 69, 72, 78)) # lower 48 only
```

## Graphs 
### Examining drought using a monthly average of DSCI  
I will begin by examining how different states experience drought across the year, using data from 2000 to 2022:  
```{r graph_monthlyDSCIaverage_byState_fill, include = TRUE, message = FALSE, warning = FALSE, fig.width=6, fig.height=10}
drought_fips_monthlyAverage_state  %>% 
  ggplot(aes(x = Month, y = State, fill = Avg_DSCI)) +
  geom_tile()+
  scale_fill_viridis_c(option = "B", begin = 0.15, end = 0.98, name = "Average DSCI")+
  scale_y_discrete(limits = rev) +
  coord_cartesian(expand = F) +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.line = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = "Drought Severity and Coverage Index (DSCI) monthly average (2000-2022)",
       subtitle = "Drought severity peaks at different times of the year for different states")

```
Visualised over time here:  
```{r graph_monthlyDSCIaverage_byState_density, include = TRUE, message = FALSE, warning = FALSE}
US_timelapse_DSCI_byMonth <-drought_fips_orig %>% 
  mutate(Year = year(date),
         Month = month(date, label = TRUE)) %>% 
  group_by(State, Year, Month) %>% 
  summarise(Avg_DSCI = mean(DSCI, na.rm = TRUE)) %>% 
  ggplot(aes(x = Month, y = Avg_DSCI, colour = Avg_DSCI)) +
  geom_violin(color = "transparent", fill = "gray90") +
  geom_sina(size = 0.95) +
  scale_color_viridis_c(option = "B", begin = 0.15, end = 0.98) +
  coord_cartesian(expand = F) +
  theme_bw() +
  theme(axis.title.x = element_blank(),
        axis.line = element_blank(),
        axis.ticks = element_blank(),
        legend.position = "none") +
  transition_states(Year) +
  labs(title = "Drought Severity and Coverage Index (DSCI) monthly average",
       subtitle = "Year: {closest_state}",
       y = "Average DSCI")

animate(US_timelapse_DSCI_byMonth,renderer = gifski_renderer(), duration = 30, height = 350, width = 600, units = "px")
```
### Examining drought using an annualised average of DSCI  
This approach will demonstrate a broad yearly trend.  

Here, I look at drought severity in the US as a whole,  
```{r graph_annualDSCIaverage, include = TRUE, message = FALSE, warning = FALSE}
drought_fips_orig %>% 
  mutate(Year = year(date)) %>% 
  group_by(Year) %>% 
  summarise(Avg_DSCI = mean(DSCI, na.rm = TRUE)) %>% 
  ggplot(aes(x = Year, y = Avg_DSCI)) +
  geom_line(colour = "grey")+
  geom_point(aes(group = seq_along(Year), colour = Avg_DSCI), size = 2) +
  scale_colour_viridis_b()+
  scale_x_continuous(minor_breaks = seq(2000, 2022, 1)) +
  theme_classic() +
  theme(axis.title.x = element_blank(),
        panel.grid.minor.x = element_line(),
        panel.grid.major.x = element_line()) +
  labs(title = "Drought Severity and Coverage Index (DSCI) annual US average")

```
and drought severity by State here:  
```{r graph_USmap_timelapse_DSCI_byState, include = TRUE, message = FALSE, warning = FALSE}
#create a US map and plot average DSCI 2000-2022
USmap_timelapse_DSCI_byState <- states %>% 
  left_join(drought_fips_yearlyAverage_state, by = c("NAME" = "state_name")) %>% 
  ggplot(aes(fill = Avg_DSCI, group = seq_along(State))) + 
  geom_sf() +
  scale_fill_continuous(type = "viridis") +
  theme_bw() +
  transition_states(Year, transition_length = 1, state_length = 2) +
  labs(title = "Drought Severity and Coverage Index (DSCI) annual average by US State",
       subtitle = "Year: {closest_state}")

animate(USmap_timelapse_DSCI_byState, renderer = gifski_renderer(), duration = 20)
```

and here:  
```{r graph_annualDSCIaverage_byState, include = TRUE, message = FALSE, warning = FALSE}
US_timelapse_DSCI_byState <- drought_fips_yearlyAverage_state %>% 
  ggplot(aes(x = Avg_DSCI, y = state_name)) +
  geom_point(aes(group = seq_along(State), colour = Avg_DSCI), size = 2) +
  scale_colour_viridis_b()+
  scale_y_discrete(limits=rev) +
  theme_bw() +
  theme(axis.title.y = element_blank()) +
  transition_states(Year) +
  labs(title = "Drought Severity and Coverage Index (DSCI) annual average by US State",
       subtitle = "Year: {closest_state}") +
  ease_aes('cubic-in-out')

animate(US_timelapse_DSCI_byState,renderer = gifski_renderer(), duration = 20, height = 650, width = 600, units = "px")
```






# Key Takeaways  
- Drought intensity varies greatly by state and year
- California has experienced some of the most intense droughts since 2000  
- In some years, drought is widespread. In other years, it is more localised.
- The average annual drought intensity across the US was highest in 2012 and lowest in 2019

# Possible next steps  
- Examine/visualise data at monthly average drought intensity rather than current annual average
- Integration of population data - where are the most people affected by drought?  