---
title: "TidyTuesday Week 19: NYT Bestsellers"
author: "Tim Kelso"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE)
```

```{r libraries}
library(tidytuesdayR)
library(DataExplorer)
library(lubridate)
library(tidyverse)
```
# Background 
## Data source  
Data from Post45 Data (https://data.post45.org/wp-content/uploads/2022/01/NYT-Data-Description.pdf)
## Data information  
Fiction bestseller list between 1931 and 2020.
Each row of the dataset is a single "entry" on the list - single slot for a single week

*nyt_full.tsv* 
- year: the year of appearance  
- week: the weekly issue of the bestseller list 
- rank: the book's rank on the list for that week  
- title_id: a unique ID mapping titles to the nyt_titles spreadsheet  
- title: title of the novel, as reported by the New York Times  
- author: author of the novel, as reported by the New York Times  

*nyt_titles.tsv*
- id: an arbitrary unique id for the novel  
- title: the title of the novel, as reported by the New York Times  
- author: the author of the novel, as reported by the New York Times  
- year: the first year that the novel appears on the bestseller list. Note that this year may be different from the publication year.  
- total_weeks: the total number of weeks the title was on the list 
- first_week: the first week that the novel appears on the bestseller list 
- debut_rank: the book's bestseller rank in the week of its first appearance
- best_rank: the highest rank achieved by the title while on the list  

```{r get_data}
# Get the Data
nyt_titles_orig <- readr::read_tsv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-05-10/nyt_titles.tsv')

nyt_full_orig <- readr::read_tsv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-05-10/nyt_full.tsv')

```

```{r inspect_data}
head(nyt_full_orig)
```

# Analysis  
##Initial thoughts  
- What is the average title length? Does this change over time?   
- Do shorter/longer titles stay longer on the list?    
- Who is the most prolific author since 1931?  
-   


## Tables/Columns of potential interest:  
- Line graph - average title length by year (nyt_full)  
- total weeks by title length (nyt_titles)  
- Wordcloud of author frequency in total weeks (nyt_full)  
  


```{r data_transformation}

```

## Graphs  
```{r graph1_, include = TRUE, message = FALSE}

```


```{r graph2_, include = TRUE, message = FALSE}

```

# Key Takeaways  
-  
-  
-  
- 

# Possible next steps  
-  
-  
-  
-  