---
title: "Solar/Wind utilities in USA: TidyTuesday Week 18, 2022"
author: "Tim Kelso"  
date: "`r format(Sys.time(), '%d %B, %Y')`"  
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE)
```

```{r libraries}
library(tidytuesdayR)
library(DataExplorer)
library(tidytext)
library(lubridate)
library(tidyverse)
```
# Background 
## Data source
https://emp.lbl.gov/utility-scale-solar

## Data information
Berkeley Lab’s “Utility-Scale Solar, 2021 Edition” presents analysis of empirical plant-level data from the U.S. fleet of ground-mounted photovoltaic (PV), PV+battery, and concentrating solar-thermal power (CSP) plants with capacities exceeding 5 MWAC. While focused on key developments in 2020, this report explores trends in deployment, technology, capital and operating costs, capacity factors, the levelized cost of solar energy (LCOE), power purchase agreement (PPA) prices, and wholesale market value.


```{r get_data}
# Get the Data
tuesdata <- tidytuesdayR::tt_load(2022, week = 18)

capacity_orig <- tuesdata$capacity
wind_orig <- tuesdata$wind
solar_orig <- tuesdata$solar
average_cost_orig <- tuesdata$average_cost
```

# Initial thoughts  
## Data  
- Capacity: Time series data of power generation capacity, from 2014-2020, for solar, wind, storage, gas, nuclear, coal, other. _prior and _new are cumulative totals?  
- Average_cost: Average cost of gas, solar, wind per mWh from 2014-2020 (table needs pivot_longer)  
- Wind: Cost and capacity, with a more granular time scale  
- Solar: Cost and capacity, with a more granular time scale  

## Key questions  
1. How has capacity and average cost of energy generation changed from 2014-2020?  
2. Is there a correlation between the capacity of energy sources (wind, solar, gas) and average cost?  

# Analysis  
## Graphs of potential interest  
- Simple line graph of gas vs solar vs wind total capacity from 2014-2020 (capacity)  
- Simple line graph of gas vs solar vs wind average_cost from 2014-2020 (average_cost)  
- Comparison of average cost vs capacity for gas, solar, wind (capacity, average_cost)  
- Standalone prior vs standalone new over time for solar, wind, storage, gas, nuclear, coal, other (capacity)  
- Hybrid prior vs hybrid new for solar, wind, storage, gas, nuclear, coal, other (capacity)  


## Graphs
```{r data_transformation}
capacity <- capacity_orig %>% 
  pivot_longer(cols = c(standalone_prior:total_gw), names_to = "measure", values_to = "Capacity(GW)") %>% 
  mutate(measure = as.factor(measure))

average_cost <- average_cost_orig %>% 
  rename(gas = gas_mwh, solar = solar_mwh, wind = wind_mwh) %>% 
  pivot_longer(cols = c(gas:wind), names_to = "source", values_to = "dollars_per_MWh") %>% 
  mutate(source = as.factor(source))
```

```{r plot_capacity_over_time, include = TRUE, message = FALSE}
capacity %>% 
  filter(measure == "total_gw") %>% 
  ggplot(aes(x = year, y = `Capacity(GW)`, grouping = type)) +
  theme_classic() +
  labs(title = "Energy generation capacity in the USA",
       x = element_blank()) +
  geom_line(aes(colour = type), size = 1) + 
  scale_colour_viridis_d("Power Type", option = "magma") +
  coord_cartesian(expand = FALSE) +
  geom_label(x = 2019, y = 425, color = "#FF9999", size = 3.5, label = "Solar capacity has risen steadily since 2015", hjust = 1)
```
```{r plot_cost_over_time, include = TRUE, message = FALSE}
average_cost %>% 
  ggplot(aes(x = year, y = dollars_per_MWh,grouping = source)) +
  theme_classic() +
  labs(title = "Average cost of generating power ($/MWh) in the USA",
       x = element_blank(),
       y = "$/MWh") +
  geom_line(aes(colour = source), size = 1) + 
  scale_colour_viridis_d("Power Type", option = "inferno") +
  scale_x_continuous(breaks=seq(2009, 2021, 2)) +
  coord_cartesian(expand = FALSE) +
  geom_label(x = 2010, y = 153, color = "#FF9999", size = 3.5, label = str_wrap("Cost of generating solar power has dropped dramatically since 2009 and is now lower than gas and wind", 50), hjust = 0)
```

# Key Takeaways  
-  Solar power capacity has risen sharply against all other forms since 2014, whilst it's average cost has dropped dramatically since 2009 and is now lower than gas and wind  
- Wind power capacity has stagnated since 2018, matched by an increased cost in generating this type of energy during this period
-  Cost of gas power generation has decreased since 2009, despite reduced gas power capacity since 2014  

# Possible next steps  
- Improve visuals of graphs - colours, font, label doesn't really stand out  
- Visualise capacity vs average cost for solar, wind, and gas  
- Take a deeper look at recent solar and wind trends using more detailed data from individual days/months