---
title: "TidyTuesday: 21Apr2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE)
```

```{r libraries}
library(tidytuesdayR)
library(lubridate)
library(tidyverse)
```
Data source: https://towardsdatascience.com/the-wild-world-of-crossword-data-71d560e222f5

```{r get_data}
# Get the Data
tuesdata <- tidytuesdayR::tt_load('2022-04-19')

big_dave <- tuesdata$big_dave
big_dave_orig <- big_dave

times <- tuesdata$times
times_orig <- times
```

```{r inspect_data}
head(big_dave)
```

```{r data_transformation}
big_dave <- big_dave_orig %>% 
  mutate(clue_Direction = as.factor(case_when(
    str_detect(clue_number, pattern = "a") ~ "Across",
    str_detect(clue_number, pattern = "d") ~ "Down")),
  puzzle_day = wday(puzzle_date, label = T),
  puzzle_year = year(puzzle_date),
  answer_character_count = str_length(answer),
  clue_number_only = str_extract(clue_number, "[:digit:]*"),
  clue_text_only = str_extract(clue, ".*(?=[:blank:])"),
  clue_character_count = str_length(clue_text_only)
  )

```

```{r initial_graphs, include = TRUE, message = FALSE}
#compare average clue length on each weekday over the years
big_dave %>% 
  group_by(puzzle_year, puzzle_day) %>% 
  summarise(mean_clue_character_count = mean(clue_character_count, na.rm = TRUE)) %>% 
  ggplot(aes(x = puzzle_day, y = mean_clue_character_count, group = as.factor(puzzle_year))) + 
  geom_col(aes(fill = as.factor(puzzle_year)), position = "dodge") +
  scale_colour_viridis_d()+
  theme_classic()+
  coord_cartesian(expand = FALSE, ylim = c(30,50))+
  labs(title = "Daily Telegraph Crossword: Clue length (2009-2022)",
       y = "Average Clue Length (Characters)",
       x = element_blank(),
       fill = "")
```

# Key Takeaways  
-Clue length appears to increase from Monday to Saturday, then drops down on Sunday   
-Monday and Saturday clues have become markedly longer whilst Sunday clues have become much shorter  

# Possible next steps  
-Use a better colour scheme  
-Compare to New York Times Crossword data  
-Compare fewer years (reduce noise)  
-Examine weekly trends by year, for instance with line graphs  
-Predict changes for upcoming years