---
title: "TidyTuesday: 27Apr2022"
author: "Tim Kelso"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE)
```

```{r libraries}
library(tidytuesdayR)
library(DataExplorer)
library(wordcloud2)
library(tidytext)
library(lubridate)
library(tidyverse)
```
Data source: https://www.kaggle.com/code/headsortails/hidden-gems-a-collection-of-underrated-notebooks/report

The file of interest is kaggle_hidden_gems.csv, containing the following columns:

vol and date: The consecutive number of the Hidden Gems episode and when it was first published.

link_forum and link_twitter: The hyperlinks to the Kaggle Forum post and Twitter post for the episode.

notebook and author: The hyperlinks to the Notebook itself, as well as to the Kaggle profile of the author.

title: The Notebook title as a string.

review: My brief review of the Notebook.

author_name: The name of the Notebook author as listed on their Kaggle profile at the time the episode was published.

author_twitter and author_linkedin: The social media links of the author, if listed on their Kaggle profile.

notes: Notes about special episodes. 


```{r get_data}
# Get the Data
tuesdata <- tidytuesdayR::tt_load(2022, week = 17)

hidden_gems <- tuesdata$hidden_gems

hidden_gems <- tuesdata$hidden_gems
hidden_gems_orig <- hidden_gems
```

```{r inspect_data}
head(hidden_gems)
```
##Columns of potential interest:  
- author_kaggle - are there many authors with multiple Hidden Gems entries?  
- Title - how many titles are structured as questions?  
- Title - what are the most frequent words included in titles?  
- Title - what are the most frequent countries included in titles?  
- Review - what are the most frequent words included in reviews?  
- Author_twitter - how many authors have linked their twitter accounts?  
- Author_linkedin - how many authors have linked their linkedin accounts?  
- Author_linkedin & _twitter - how many authors have both, and what is the overlap?
```{r data_transformation}
#create a wordcloud of Titles
#source: https://towardsdatascience.com/create-a-word-cloud-with-r-bde3e7422e8a 
hidden_gems_title_words <- hidden_gems_orig %>%
  select(title) %>%
  unnest_tokens(word, title) %>% 
  anti_join(stop_words, by = "word") %>% 
  count(word, sort = TRUE)
```

```{r initial_graphs, include = TRUE, message = FALSE}
#most frequently used words in Titles
wordcloud2(data=hidden_gems_title_words, size=1.6, color='random-dark')

```

# Key Takeaways  
-Clue length appears to increase from Monday to Saturday, then drops down on Sunday   
-Monday and Saturday clues have become markedly longer whilst Sunday clues have become much shorter  

# Possible next steps  
-Use a better colour scheme  
-Compare to New York Times Crossword data  
-Compare fewer years (reduce noise)  
-Examine weekly trends by year, for instance with line graphs  
-Predict changes for upcoming years