---
title: "TidyTuesday: 27Apr2022"
author: "Tim Kelso"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE)
```

```{r libraries}
library(tidytuesdayR)
library(DataExplorer)
library(wordcloud2)
library(tidytext)
library(lubridate)
library(tidyverse)
```
# Background 
## Data source
https://www.kaggle.com/code/headsortails/hidden-gems-a-collection-of-underrated-notebooks/report

## Data information
The file of interest is kaggle_hidden_gems.csv, containing the following columns:

vol and date: The consecutive number of the Hidden Gems episode and when it was first published.

link_forum and link_twitter: The hyperlinks to the Kaggle Forum post and Twitter post for the episode.

notebook and author: The hyperlinks to the Notebook itself, as well as to the Kaggle profile of the author.

title: The Notebook title as a string.

review: My brief review of the Notebook.

author_name: The name of the Notebook author as listed on their Kaggle profile at the time the episode was published.

author_twitter and author_linkedin: The social media links of the author, if listed on their Kaggle profile.

notes: Notes about special episodes.  


```{r get_data}
# Get the Data
tuesdata <- tidytuesdayR::tt_load(2022, week = 17)

hidden_gems <- tuesdata$hidden_gems

hidden_gems <- tuesdata$hidden_gems
hidden_gems_orig <- hidden_gems
```

```{r inspect_data}
head(hidden_gems)
```
# Analysis  
## Columns of potential interest:  
- author_kaggle - are there many authors with multiple Hidden Gems entries?  
- Title - how many titles are structured as questions?  
- Title - what are the most frequent words included in titles?  
- Title - what are the most frequent countries included in titles?  
- Review - what are the most frequent words included in reviews?  
- Author_twitter - how many authors have linked their twitter accounts?  
- Author_linkedin - how many authors have linked their linkedin accounts?  
- Author_linkedin & _twitter - how many authors have both, and what is the overlap?  
```{r data_transformation}
#create a wordcloud of Titles
#source: https://towardsdatascience.com/create-a-word-cloud-with-r-bde3e7422e8a 
hidden_gems_title_words <- hidden_gems_orig %>%
  select(title) %>%
  unnest_tokens(word, title) %>% 
  anti_join(stop_words, by = "word") %>% 
  count(word, sort = TRUE)

#create a wordcloud of Review
hidden_gems_review_words <- hidden_gems_orig %>%
  select(review) %>%
  mutate(review = str_replace_all(review, "https", "")) %>% 
  unnest_tokens(word, review) %>% 
  anti_join(stop_words, by = "word") %>% 
  count(word, sort = TRUE)

#create wordcloud of most common authors
hidden_gems_authors_count <- hidden_gems_orig %>%
  select(author_kaggle) %>%
  count(author_kaggle, sort = TRUE)

#how many titles are structured as questions?  ie. contain a question mark
hidden_gems_title_questions <- hidden_gems_orig %>% 
  select(title) %>% 
  mutate(question = case_when(
    str_detect(title, "\\?") ~ 1,
    TRUE ~ 0
  )) %>% 
  summarise(question_count = sum(question),
            total_titles = n()) %>% 
  mutate(questionCount_proportion = round(100*question_count/total_titles, 1)) %>% 
  pivot_longer(cols = )
```

```{r titles_wordcloud, include = TRUE, message = FALSE}
#most frequently used words in Titles
wordcloud2(data=hidden_gems_title_words, color='random-dark')
```

```{r question_titles_table, include = TRUE, message = FALSE}
knitr::kable(hidden_gems_title_questions, col.names = c("Titles with questions", "Total number of Titles", "Proportion of Titles that included Questions (%) "), align = "l")
```


```{r reviews_wordcloud, include = TRUE, message = FALSE}
#most frequently used words in Titles
wordcloud2(data=hidden_gems_review_words, color='random-dark')
```

```{r authors_wordcloud, include = TRUE, message = FALSE}
#most frequently featured Authors
wordcloud2(data=hidden_gems_authors_count, size = 0.5, color='random-light')
```

# Key Takeaways  
-Lots of Hidden Gems titles focus on analysis, data, and exploratory data analysis (eda), which is not a huge shock! Other frequently used words include visualisation, world, learning, science, and time.  
-Most titles (89%) did not include a question, in line with the predominance of exploratory data analysis that would not lend itself to a specific focus in such a form  
-Reviews contain many mentions of data and visuals, with notebook and kaggle also featuring high up the list. Next on the list are detailed, survey, and analysis.  
-The most featured Author in Hidden Gems is jonathonbouchet with 9 reviews, followed by thedatabeast (5 reviews). Next on the list are lots of Authors with 3 reviews, with an honourable mention to captcalculator for an amazing kaggle username.  

# Possible next steps  
-Title - what are the most frequent countries included in titles?  
-Author_twitter - how many authors have linked their twitter accounts?  
-Author_linkedin - how many authors have linked their linkedin accounts?  
-Author_linkedin & _twitter - how many authors have both, and what is the overlap?