---
title: "TidyTuesday: 27Apr2022"
author: "Tim Kelso"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE)
```

```{r libraries}
library(tidytuesdayR)
library(DataExplorer)
library(lubridate)
library(tidyverse)
```
Data source: https://www.kaggle.com/code/headsortails/hidden-gems-a-collection-of-underrated-notebooks/report

```{r get_data}
# Get the Data
tuesdata <- tidytuesdayR::tt_load(2022, week = 17)

hidden_gems <- tuesdata$hidden_gems

hidden_gems <- tuesdata$hidden_gems
hidden_gems_orig <- hidden_gems
```

```{r inspect_data}
head(hidden_gems)
```

```{r data_transformation}

```

```{r initial_graphs, include = TRUE, message = FALSE}
#comparison:
```

# Key Takeaways  
-Clue length appears to increase from Monday to Saturday, then drops down on Sunday   
-Monday and Saturday clues have become markedly longer whilst Sunday clues have become much shorter  

# Possible next steps  
-Use a better colour scheme  
-Compare to New York Times Crossword data  
-Compare fewer years (reduce noise)  
-Examine weekly trends by year, for instance with line graphs  
-Predict changes for upcoming years