---
title: "TidyTuesday Week 19: NYT Bestsellers"
author: "Tim Kelso"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE)
```

```{r libraries}
library(tidytuesdayR)
library(extrafont)
library(DataExplorer)
library(ggwordcloud)
library(tidytext)
library(lubridate)
library(tidyverse)
```
# Background 
## Data source  
Data from Post45 Data (https://data.post45.org/wp-content/uploads/2022/01/NYT-Data-Description.pdf)
## Data information  
Fiction bestseller list between 1931 and 2020.
Each row of the dataset is a single "entry" on the list - single slot for a single week

*nyt_full.tsv* 
- year: the year of appearance  
- week: the weekly issue of the bestseller list 
- rank: the book's rank on the list for that week  
- title_id: a unique ID mapping titles to the nyt_titles spreadsheet  
- title: title of the novel, as reported by the New York Times  
- author: author of the novel, as reported by the New York Times  

*nyt_titles.tsv*
- id: an arbitrary unique id for the novel  
- title: the title of the novel, as reported by the New York Times  
- author: the author of the novel, as reported by the New York Times  
- year: the first year that the novel appears on the bestseller list. Note that this year may be different from the publication year.  
- total_weeks: the total number of weeks the title was on the list 
- first_week: the first week that the novel appears on the bestseller list 
- debut_rank: the book's bestseller rank in the week of its first appearance
- best_rank: the highest rank achieved by the title while on the list  

```{r get_data}
# Get the Data
nyt_titles_orig <- readr::read_tsv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-05-10/nyt_titles.tsv')

nyt_full_orig <- readr::read_tsv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-05-10/nyt_full.tsv')
```

```{r inspect_data}
head(nyt_full_orig)
```

# Analysis  
## Initial thoughts  
- What is the average title length? Does this change over time?   
- Do shorter/longer titles stay longer on the list?    
- Who is the most prolific author since 1931?  
-   


## Tables/Columns of potential interest:  
- Line graph - average title length by year (nyt_full)  
- Total weeks by title length (nyt_titles)  
- Wordcloud of author frequency in total weeks (nyt_full)  
  


```{r data_transformation}

```

## Graphs  
```{r graph_most_common_words, include = TRUE, message = FALSE, warning = FALSE}
#create a wordcloud of Titles
nyt_titles_orig %>%
  select(title) %>%
  unnest_tokens(word, title) %>% 
  anti_join(stop_words, by = "word") %>% 
  count(word, sort = TRUE) %>% 
  head(400) %>% #take top 400
  ggplot((aes(label = word, size = n, col = n))) +
  geom_text_wordcloud(seed = 3590, family = "Tw Cen MT") + 
  labs(title = "Most Common Words in NYT Bestsellers Titles")+
  scale_color_distiller(palette="Dark2") +
  theme_void()
```

```{r graph_bestselling_authors_by_total_weeks, include = TRUE, message = FALSE}
nyt_titles_orig %>% 
  group_by(author) %>% 
  summarise(author_total_weeks = sum(total_weeks)) %>% 
  arrange(desc(author_total_weeks)) %>% 
  head(200) %>% #take top 200 ()
  ggplot((aes(label = author, size = author_total_weeks, col = author_total_weeks))) +
  geom_text_wordcloud(seed = 3590, family = "Tw Cen MT") + 
  labs(title = "Bestselling Authors by Total Number of Weeks in Bestsellers List")+
  scale_color_distiller(palette="Dark2") +
  theme_void()
```

```{r graph_title_length, include = TRUE, message = FALSE}
nyt_titles_orig %>% 
  mutate(title_length = str_length(title)) %>% 
  ggplot(aes(x = title_length)) +
  geom_histogram() +
  geom_vline(aes(xintercept = mean(title_length)),col='red',size=1)+
  scale_x_continuous(breaks = seq(0, 70, 5)) +
  scale_y_continuous(breaks = seq(0, 2400, 200)) +
  theme_classic()+
  labs(title = "NYT Bestsellers: Title Lengths (1931-2020)",
       x = "Title Length (characters)",
       y = "Number of books") +
  coord_cartesian(expand = F)
```

```{r graph_average_title_length_by_year, include = TRUE, message = FALSE}
nyt_titles_orig %>% 
  mutate(title_length = str_length(title)) %>% 
  group_by(year) %>% 
  summarise(average_title_length = mean(title_length)) %>% 
  ggplot(aes(x = year, y = average_title_length)) +
  geom_line() +
  geom_smooth(method = "lm") +
  theme_classic() +
  scale_x_continuous(breaks = seq(1930, 2020, 10)) +
  labs(title = "NYT Bestsellers: Average Title Length (1931-2020)",
       x = element_blank(),
       y = "Average Title Length (characters)")
```

```{r graph_title_length_by_total_weeks, include = TRUE, message = FALSE}
nyt_titles_orig %>% 
  mutate(title_length = str_length(title)) %>% 
  ggplot(aes(x = title_length, y = total_weeks)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_classic() +
  labs(title = "NYT Bestsellers: Does Title Length Correlate with Time in Bestseller List?",
       x = "Title Length (characters)",
       y = "Time in Bestseller List (weeks)")
```



# Key Takeaways  
- Average title length of NYT bestseller books is ~15 characters (1931-2020)  
- Average title length is on a downward trend since 1931  
- There is not a strong relationship between title length and total weeks in the bestseller list  
- Books by Danielle Steel, Stephen King, and John Grisham have spent the longest time in the bestseller list  
- "Night","Love", "Death", "House", "Dark", and "Time", have appeared most often in bestseller titles since 1931

# Possible next steps  
- Do the top 200 authors have a few books with a long stay in the bestseller list, or many books with medium/low time in list?  
- Are there any other characteristics that determine time spent in the list? For instance, does a high debut rank predict longer time spent in the list?  