---
title: "2023, Week 15: Regionality of new and disestablished schools"
author: "Tim Kelso"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE)

library(lubridate)
library(readxl)
library(tidyverse)
```
# Background 
## Data source  
Data from Australian Curriculum, Assessment and Reporting Authority (ACARA) (https://acara.edu.au/contact-us/acara-data-access).  

## Data information  
*ACARA_school_location_.xlsx*
List of all Australian schools, Long/Lat, LGA

*ACARA_school_profile_.xlsx*
List of all Australian schools, enrolments, ICSEA, LBOTE, SEA


```{r get_data}
# Get the Data
schoolProfile_yearly <- read_excel("ACARA_school_profile_2008-2022.xlsx", sheet = "SchoolProfile 2008-2022", .name_repair = "universal")
schoolLocation_2022 <- read_excel("ACARA_school_location_2022.xlsx", sheet = "SchoolLocations 2022", .name_repair = "universal") %>% mutate(Calendar.Year = 2022)
schoolLocation_2021 <- read_excel("ACARA_school_location_2021.xlsx", sheet = "SchoolLocations 2021", .name_repair = "universal")%>% mutate(Calendar.Year = 2021)
schoolLocation_2020 <- read_excel("ACARA_school_location_2020.xlsx", sheet = "SchoolLocations 2020", .name_repair = "universal")%>% mutate(Calendar.Year = 2020)
schoolLocation_2019 <- read_excel("ACARA_school_location_2019.xlsx", sheet = "School Location 2019", .name_repair = "universal")%>% mutate(Calendar.Year = 2019)
schoolProfile_2022 <- read_excel("ACARA_school_profile_2022.xlsx", sheet = "SchoolProfile 2022", .name_repair = "universal")
schoolProfile_2021 <- read_excel("ACARA_school_profile_2021.xlsx", sheet = "SchoolProfile 2021", .name_repair = "universal")
```

# Analysis  
## Initial thoughts  
- What state has gained the most/least number of schools since 2019?  
- How many schools were gained in major cities vs regional areas?
- How many schools were lost in major cities vs regional areas?

## Columns of potential interest:  
- Geolocation    
- ICSEA  

```{r data_prep}
#create school_location list combining data from 2019-2022
schoolLocation_yearly <- schoolLocation_2019  %>%
  bind_rows(schoolLocation_2020, schoolLocation_2021, schoolLocation_2022) %>% 
  mutate(across(.cols = c(State, Campus.Type:Rolled.Reporting.Description, ABS.Remoteness.Area.Name), .fns = as.factor)) %>% 
  mutate(ABS.Remoteness.Area.Name = factor(ABS.Remoteness.Area.Name, levels = c("Major Cities", "Inner Regional", "Outer Regional", "Remote", "Very Remote"), ordered = TRUE),
         School.Sector = factor(School.Sector, levels = c("Government", "Catholic", "Independent")),
         School.Type = factor(School.Type, levels = c("Primary", "Secondary", "Combined", "Special")))

schoolProfile_yearly <- schoolProfile_yearly  %>%
  mutate(across(.cols = c(State, Campus.Type:Rolled.Reporting.Description, Governing.Body, Year.Range), .fns = as.factor)) %>% 
  mutate(Geolocation = factor(Geolocation, levels = c("Major Cities", "Inner Regional", "Outer Regional", "Remote", "Very Remote"), ordered = TRUE),
         School.Sector = factor(School.Sector, levels = c("Government", "Catholic", "Independent")),
         School.Type = factor(School.Type, levels = c("Primary", "Secondary", "Combined", "Special")))
```
## Graphs 
### Total number of schools in Australia 2019-2022  
I'll start with a simple count of total schools in Australia using school_location data from 2019 to 2022:  

``` {r uniqueSchools, include = TRUE, message = FALSE}
schoolLocation_yearly %>% 
  group_by(Calendar.Year) %>% 
  summarise(n = n()) %>% 
  pivot_wider(names_from = Calendar.Year, values_from = n)
```

### Gained and lost schools 2019-2022
The above shows the final number of schools each year. Let's find how many schools were lost and gained in each year from 2019 onwards.

```{r gained_lost,  include = TRUE, message = FALSE}
#anti-join
schoolsGained_2020 <- schoolLocation_2020 %>% 
  anti_join(schoolLocation_2019, by = "ACARA.SML.ID") %>% 
  count() %>% 
  mutate(Calendar.Year = 2020,
         Group = "Gained")

schoolsLost_2020 <- schoolLocation_2019 %>% 
  anti_join(schoolLocation_2020, by = "ACARA.SML.ID") %>% 
  count() %>% 
  mutate(Calendar.Year = 2020,
         Group = "Lost")


schoolsGained_2021 <- schoolLocation_2021 %>% 
  anti_join(schoolLocation_2020, by = "ACARA.SML.ID") %>% 
  count() %>% 
  mutate(Calendar.Year = 2021,
         Group = "Gained")

schoolsLost_2021 <- schoolLocation_2020 %>% 
  anti_join(schoolLocation_2021, by = "ACARA.SML.ID") %>% 
  count() %>% 
  mutate(Calendar.Year = 2021,
         Group = "Lost")


schoolsGained_2022 <- schoolLocation_2022 %>% 
  anti_join(schoolLocation_2021, by = "ACARA.SML.ID") %>% 
  count() %>% 
  mutate(Calendar.Year = 2022,
         Group = "Gained")

schoolsLost_2022 <- schoolLocation_2021 %>% 
  anti_join(schoolLocation_2022, by = "ACARA.SML.ID") %>% 
  count() %>% 
  mutate(Calendar.Year = 2022,
         Group = "Lost")

schoolsGained_2020 %>% 
  bind_rows(schoolsLost_2020, schoolsGained_2021, schoolsLost_2021, schoolsGained_2022, schoolsLost_2022) %>% 
  mutate(n = if_else(Group == "Lost", -n, n)) %>% 
  ggplot(aes(x = as.factor(Calendar.Year), y = n, group = Group)) +
  geom_col(aes(fill = Group), position = "stack") +
  geom_text(aes(label = n), vjust = 0, size = 5) +
  geom_hline(yintercept = 0) +
  scale_fill_manual(values = c("#3E6990", "#F39B6D"))+
  theme_classic()+
  labs(title = "~2x the number of schools are established than lost each year",
       y = "Number of schools") +
  theme(axis.title.x = element_blank(),
        axis.text = element_text(size = 10),
        legend.title = element_blank(),
        legend.text = element_text(size = 9))
```
### Which school systems gain or lose more schools?  
```{r gained_lost_bySystem,  include = TRUE, message = FALSE, fig.height = 8}
schoolsGained_2020 <- schoolLocation_2020 %>% 
  anti_join(schoolLocation_2019, by = "ACARA.SML.ID") %>% 
  group_by(School.Sector) %>% 
  count() %>% 
  mutate(Calendar.Year = 2020,
         Group = "Gained"
         )

schoolsLost_2020 <- schoolLocation_2019 %>% 
  anti_join(schoolLocation_2020, by = "ACARA.SML.ID") %>% 
  group_by(School.Sector) %>% 
  count() %>% 
  mutate(Calendar.Year = 2020,
         Group = "Lost")


schoolsGained_2021 <- schoolLocation_2021 %>% 
  anti_join(schoolLocation_2020, by = "ACARA.SML.ID") %>% 
  group_by(School.Sector) %>%  
  count() %>% 
  mutate(Calendar.Year = 2021,
         Group = "Gained")

schoolsLost_2021 <- schoolLocation_2020 %>% 
  anti_join(schoolLocation_2021, by = "ACARA.SML.ID") %>% 
  group_by(School.Sector) %>% 
  count() %>% 
  mutate(Calendar.Year = 2021,
         Group = "Lost")


schoolsGained_2022 <- schoolLocation_2022 %>% 
  anti_join(schoolLocation_2021, by = "ACARA.SML.ID") %>% 
  group_by(School.Sector) %>% 
  count() %>% 
  mutate(Calendar.Year = 2022,
         Group = "Gained")

schoolsLost_2022 <- schoolLocation_2021 %>% 
  anti_join(schoolLocation_2022, by = "ACARA.SML.ID") %>% 
  group_by(School.Sector) %>% 
  count() %>%  
  mutate(Calendar.Year = 2022,
         Group = "Lost")

schoolsGained_2020 %>% 
  bind_rows(schoolsLost_2020, schoolsGained_2021, schoolsLost_2021, schoolsGained_2022, schoolsLost_2022) %>% 
  mutate(n = if_else(Group == "Lost", -n, n),
         School.Sector = factor(School.Sector, levels = c("Government", "Catholic", "Independent"), ordered = TRUE)) %>% 
  ggplot(aes(x = Calendar.Year, y = n, alpha = Group)) +
  geom_col(aes(fill = School.Sector), position = "stack") +
  geom_text(aes(label = n, vjust = if_else(Group == "Gained", 0, 1))) +
  geom_hline(yintercept = 0) +
  scale_fill_manual(values = c("#381D2A",	"#3E6990",	"#AABD8C")) +
  scale_alpha_discrete(range = c(1, 0.7))+
#  scale_x_discrete(position = "top") +
  theme_classic()+
  labs(title = "~50% of schools established in the last three years were Independent schools",
       y = "Number of schools gained/lost") +
  theme(axis.title.x = element_blank(),
        axis.text = element_text(size = 10),
        legend.title = element_blank(),
        legend.position = "none",
        legend.text = element_text(size = 9),
        strip.text = element_text(size = 12)) +
  facet_wrap(nrow = 1, vars(School.Sector))

ggsave("./images/Established schools by sector.png", width = 15, dpi = 300)
```

### Did major cities or regional areas gain or lose more schools?  

```{r gained_lost_byRegion_aggregated, include = TRUE, message = FALSE, dpi = 300}
#anti-join
schoolsGained_2020 <- schoolLocation_2020 %>% 
  anti_join(schoolLocation_2019, by = "ACARA.SML.ID") %>% 
  mutate(Region = case_when(
    ABS.Remoteness.Area.Name == "Major Cities" ~ "Major Cities",
    TRUE ~ "Regional"
  )) %>% 
  group_by(Region) %>% 
  count() %>% 
  mutate(Calendar.Year = 2020,
         Group = "Gained"
         )

schoolsLost_2020 <- schoolLocation_2019 %>% 
  anti_join(schoolLocation_2020, by = "ACARA.SML.ID") %>% 
  mutate(Region = case_when(
    ABS.Remoteness.Area.Name == "Major Cities" ~ "Major Cities",
    TRUE ~ "Regional"
  )) %>% 
  group_by(Region) %>% 
  count() %>% 
  mutate(Calendar.Year = 2020,
         Group = "Lost")


schoolsGained_2021 <- schoolLocation_2021 %>% 
  anti_join(schoolLocation_2020, by = "ACARA.SML.ID") %>% 
  mutate(Region = case_when(
    ABS.Remoteness.Area.Name == "Major Cities" ~ "Major Cities",
    TRUE ~ "Regional"
  )) %>% 
  group_by(Region) %>% 
  count() %>% 
  mutate(Calendar.Year = 2021,
         Group = "Gained")

schoolsLost_2021 <- schoolLocation_2020 %>% 
  anti_join(schoolLocation_2021, by = "ACARA.SML.ID") %>% 
  mutate(Region = case_when(
    ABS.Remoteness.Area.Name == "Major Cities" ~ "Major Cities",
    TRUE ~ "Regional"
  )) %>% 
  group_by(Region) %>% 
  count() %>% 
  mutate(Calendar.Year = 2021,
         Group = "Lost")


schoolsGained_2022 <- schoolLocation_2022 %>% 
  anti_join(schoolLocation_2021, by = "ACARA.SML.ID") %>% 
  mutate(Region = case_when(
    ABS.Remoteness.Area.Name == "Major Cities" ~ "Major Cities",
    TRUE ~ "Regional"
  )) %>% 
  group_by(Region) %>% 
  count() %>% 
  mutate(Calendar.Year = 2022,
         Group = "Gained")

schoolsLost_2022 <- schoolLocation_2021 %>% 
  anti_join(schoolLocation_2022, by = "ACARA.SML.ID") %>% 
  mutate(Region = case_when(
    ABS.Remoteness.Area.Name == "Major Cities" ~ "Major Cities",
    TRUE ~ "Regional"
  )) %>% 
  group_by(Region) %>% 
  count() %>%  
  mutate(Calendar.Year = 2022,
         Group = "Lost")

schoolsGained_2020 %>% 
  bind_rows(schoolsLost_2020, schoolsGained_2021, schoolsLost_2021, schoolsGained_2022, schoolsLost_2022) %>% 
  mutate(n = if_else(Group == "Lost", -n, n)) %>% 
  ggplot(aes(x = Calendar.Year, y = n, group = Group)) +
  geom_col(aes(fill = Group), position = "stack") +
  geom_text(aes(label = n), vjust = 0) +
  geom_hline(yintercept = 0) +
  scale_fill_manual(values = c("#3E6990", "#F39B6D"))+
#  scale_x_discrete(position = "top") +
  theme_classic()+
  labs(title = "2-3x more schools are established each year in major cities than regional areas",
       y = "Number of schools gained/lost") +
  theme(title = element_text(face = "bold"),
        axis.title.x = element_blank(),
        axis.text = element_text(size = 10),
        legend.title = element_blank(),
        legend.text = element_text(size = 9),
        strip.text = element_text(size = 12)) +
  facet_wrap(vars(Region))

ggsave("./images/Established schools in major cities and regional areas.png", width = 8, dpi = 300)
```

### Which regional areas gain or lose more schools?  
```{r gained_lost_byRegion,  include = TRUE, message = FALSE, fig.height = 8}
schoolsGained_2020 <- schoolLocation_2020 %>% 
  anti_join(schoolLocation_2019, by = "ACARA.SML.ID") %>% 
  group_by(ABS.Remoteness.Area.Name) %>% 
  count() %>% 
  mutate(Calendar.Year = 2020,
         Group = "Gained"
         )

schoolsLost_2020 <- schoolLocation_2019 %>% 
  anti_join(schoolLocation_2020, by = "ACARA.SML.ID") %>% 
  group_by(ABS.Remoteness.Area.Name) %>% 
  count() %>% 
  mutate(Calendar.Year = 2020,
         Group = "Lost")


schoolsGained_2021 <- schoolLocation_2021 %>% 
  anti_join(schoolLocation_2020, by = "ACARA.SML.ID") %>% 
  group_by(ABS.Remoteness.Area.Name) %>%  
  count() %>% 
  mutate(Calendar.Year = 2021,
         Group = "Gained")

schoolsLost_2021 <- schoolLocation_2020 %>% 
  anti_join(schoolLocation_2021, by = "ACARA.SML.ID") %>% 
  group_by(ABS.Remoteness.Area.Name) %>% 
  count() %>% 
  mutate(Calendar.Year = 2021,
         Group = "Lost")


schoolsGained_2022 <- schoolLocation_2022 %>% 
  anti_join(schoolLocation_2021, by = "ACARA.SML.ID") %>% 
  group_by(ABS.Remoteness.Area.Name) %>% 
  count() %>% 
  mutate(Calendar.Year = 2022,
         Group = "Gained")

schoolsLost_2022 <- schoolLocation_2021 %>% 
  anti_join(schoolLocation_2022, by = "ACARA.SML.ID") %>% 
  group_by(ABS.Remoteness.Area.Name) %>% 
  count() %>%  
  mutate(Calendar.Year = 2022,
         Group = "Lost")

schoolsGained_2020 %>% 
  bind_rows(schoolsLost_2020, schoolsGained_2021, schoolsLost_2021, schoolsGained_2022, schoolsLost_2022) %>% 
  mutate(n = if_else(Group == "Lost", -n, n),
         ABS.Remoteness.Area.Name = factor(ABS.Remoteness.Area.Name, levels = c("Major Cities", "Inner Regional", "Outer Regional", "Remote", "Very Remote"), ordered = TRUE)) %>% 
  ggplot(aes(x = Calendar.Year, y = n, alpha = Group)) +
  geom_col(aes(fill = ABS.Remoteness.Area.Name), position = "stack") +
  geom_text(aes(label = n, vjust = if_else(Group == "Gained", 0, 1))) +
  geom_hline(yintercept = 0) +
  scale_alpha_discrete(range = c(1, 0.7))+
  scale_fill_manual(values = c("#090c08",	"#474056",	"#757083",	"#8a95a5",	"#b9c6ae"))+
#  scale_x_discrete(position = "top") +
  theme_classic()+
  labs(title = "Remote and very remote areas did not gain/lose many schools",
       y = "Number of schools gained/lost") +
  theme(axis.title.x = element_blank(),
        axis.text = element_text(size = 10),
        legend.title = element_blank(),
        legend.position = "none",
        legend.text = element_text(size = 9),
        strip.text = element_text(size = 12)) +
  facet_wrap(nrow = 1, vars(ABS.Remoteness.Area.Name))

ggsave("./images/Established schools in all areas.png", width = 15, dpi = 300)
```



# Key Takeaways  
-	Most new schools are established in major cities, with a net growth in the total number of schools  
- Some schools are established in regional areas, but this is offset by an equivalent number of schools that are lost  


# Possible next steps  
