---
title: "2023, Week 15: Regionality of new and disestablished schools"
author: "Tim Kelso"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE)
```

```{r libraries}
library(lubridate)
library(readxl)
library(tidyverse)
```
# Background 
## Data source  
Data from Australian Curriculum, Assessment and Reporting Authority (ACARA) (https://acara.edu.au/contact-us/acara-data-access).  

## Data information  
*ACARA_school_location_*.xlsx*
List of all Australian schools, Long/Lat, LGA

*ACARA_school_profile_*.xlsx*
List of all Australian schools, enrolments, ICSEA, LBOTE, SEA


```{r get_data}
# Get the Data
schoolProfile_yearly <- read_excel("ACARA_school_profile_2008-2022.xlsx", sheet = "SchoolProfile 2008-2022", .name_repair = "universal")
schoolLocation_2022 <- read_excel("ACARA_school_location_2022.xlsx", sheet = "SchoolLocations 2022", .name_repair = "universal") %>% mutate(Calendar.Year = 2022)
schoolLocation_2021 <- read_excel("ACARA_school_location_2021.xlsx", sheet = "SchoolLocations 2021", .name_repair = "universal")%>% mutate(Calendar.Year = 2021)
schoolLocation_2020 <- read_excel("ACARA_school_location_2020.xlsx", sheet = "SchoolLocations 2020", .name_repair = "universal")%>% mutate(Calendar.Year = 2020)
schoolLocation_2019 <- read_excel("ACARA_school_location_2019.xlsx", sheet = "School Location 2019", .name_repair = "universal")%>% mutate(Calendar.Year = 2019)
schoolProfile_2022 <- read_excel("ACARA_school_profile_2022.xlsx", sheet = "SchoolProfile 2022", .name_repair = "universal")
schoolProfile_2021 <- read_excel("ACARA_school_profile_2021.xlsx", sheet = "SchoolProfile 2021", .name_repair = "universal")
```

# Analysis  
## Initial thoughts  
- What state has gained the most/least number of schools since 2019?  
- How many schools were gained in major cities vs regional areas?
- How many schools were lost in major cities vs regional areas?

## Columns of potential interest:  
- Geolocation    
- ICSEA  

```{r data_prep}
#create school_location list combining data from 2019-2022
schoolLocation_yearly <- schoolLocation_2019  %>%
  bind_rows(schoolLocation_2020, schoolLocation_2021, schoolLocation_2022) %>% 
  mutate(across(.cols = c(State, Campus.Type:Rolled.Reporting.Description, ABS.Remoteness.Area.Name), .fns = as.factor)) %>% 
  mutate(School.Sector = factor(School.Sector, levels = c("Government", "Catholic", "Independent")),
         School.Type = factor(School.Type, levels = c("Primary", "Secondary", "Combined", "Special")))

schoolProfile_yearly <- schoolProfile_yearly  %>%
  mutate(across(.cols = c(State, Campus.Type:Rolled.Reporting.Description, Governing.Body, Year.Range), .fns = as.factor)) %>% 
  mutate(Geolocation = factor(Geolocation, levels = c("Major Cities", "Inner Regional", "Outer Regional", "Remote", "Very Remote"), ordered = TRUE),
         School.Sector = factor(School.Sector, levels = c("Government", "Catholic", "Independent")),
         School.Type = factor(School.Type, levels = c("Primary", "Secondary", "Combined", "Special")))
```
## Graphs 
### Examining drought using a monthly average of DSCI  
I'll start by examining how different states gained/lost schools using school_location data from 2019 to 2022:  

``` {r uniqueSchools}
schoolLocation_yearly %>% 
  group_by(Calendar.Year) %>% 
  summarise(n = n())
  #arrange(ACARA.SML.ID, Calendar.Year) %>% 
  #mutate(index = rank(Calendar.Year, ties.method = "first"),
  #       maxIndex = max(index),
  #       firstYear = min(Calendar.Year),
  #       finalYear = max(Calendar.Year),
  #       schoolOpen = ifelse(finalYear >= Calendar.Year, "Yes", "No"),
  #       status = case_when(
   #        index == 1 ~ "New",
   #        index >= 2 ~ "Existing"
  #       )) %>% 
  #ungroup() %>% 
  #count(ACARA.SML.ID, firstYear, finalYear)
#  filter(maxIndex < 14, firstYear == 2008)
  #group_by(Calendar.Year, status) %>% 
  #summarise(n = n()) %>% 
  ungroup() %>% 
#i suspect that schools that are not in existence in 2021 are not included in this list - it seems uncanny that all schools in this list have their final year as 2021, indicating that no schools have closed permanently since the list started
  ggplot(aes(x = Calendar.Year, y = n, grouping = status)) +
  geom_col(aes(fill = status)) +
  geom_text(aes(label = n), size = 2.5, vjust = -0.5)+
  theme_classic() +
  labs(title = "Number of Australian Schools by Year",
       y = "Count") +
  theme(axis.title.x = element_blank()) + 
  #scale_x_continuous(breaks = c(2008, 2010, 2012, 2014, 2016, 2018, 2020))+
  ylim(0, 10000)+
  coord_cartesian(expand = FALSE)
```


```{r comparison of profile and location lists}
schoolLocation_2021 %>% 
  summarise(n = n())

schoolProfile_2021 %>% 
  summarise(n = n())

schoolsInLocationListOnly <- schoolLocation_2021 %>% 
  anti_join(schoolList_profile_2021, by = "ACARA.SML.ID") 

#comparison of ACARAs not in dynamics with schools only on location list
ACARAs_NotInDynamics %>% 
  anti_join(schoolsInLocationListOnly, by = c("Acara" = "ACARA.SML.ID")) %>% 
  summarise(n = n())

#write.csv(schoolsInLocationListOnly, "../../QTA/data_raw/ACARAList_ToLoadInDynamics.csv")
```

# Key Takeaways  
- Drought intensity varies greatly by state and year
- California has experienced some of the most intense droughts since 2000  
- In some years, drought is widespread. In other years, it is more localised.
- The average annual drought intensity across the US was highest in 2012 and lowest in 2019

# Possible next steps  
- Examine/visualise data at monthly average drought intensity rather than current annual average
- Integration of population data - where are the most people affected by drought?  
